---
title: "RAD-Median and Zt-Mean"
author: "Isaac Rankin"
date: "2024"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list = ls())
library(ggplot2)
```



# Function to Compute the RAD-Median and Trimmed Mean

```{r function}
TRIMMEDmean = function(X, RAD_Median = TRUE, TRIM_percentage = 80) {
  # Set RAD_Median = TRUE to compute the RAD-Median
  # To compute the Zt-Mean, set RAD_Median = FALSE and choose the percentage of points to trim using TRIM_percentage
  
  # X is a matrix of point coordinates
  n = nrow(X) # number of observations 
  p = ncol(X) # dimension of data and must be greater or equal to 2 
  rownames(X) = paste0(rep("x", n), 1:n) # xi is observation 1 
  colnames(X) = paste0(rep("c", p), 1:p) # ci is coordinate i 
  
  # If RAD_Median = TRUE, use TRIM_Remain = 1
  if (RAD_Median == TRUE) {
    TRIM_Remain = 1
  }
  
  # If RAD_Median = FALSE, use TRIM_proportion and n to find TRIM_Remain
  if (RAD_Median == FALSE) {
    TRIM_Remain = min(n, max(1, round(n * (1 - TRIM_percentage/100))))
  }
  
  # If all points are kept, return the regular mean 
  if (TRIM_Remain == n) {
    print("The mean of all points is:")
    return(colMeans(X)) 
  } 

  # Matrix of pairwise distances
  D = as.matrix(dist(X, method = "euclidean"))
  colnames(D) = paste0(rep("x", n), 1:n) 
  rownames(D) = paste0(rep("x", n), 1:n) 

  # Find points for trimmed mean or median 
  points = n # initially have n points 
  prev_points = n # number of points in previous iteration 
  POINTS_to_use = 0 

  # Iteratively remove points
  while (points > TRIM_Remain) { 
    Prev = D 
    prev_points = ncol(Prev) 
    
    #### Handle the single point RAD-Median case separately 
    if (points == 3 & TRIM_Remain == 1) { 
      three_points = rownames(D) 
      farthest = which(D == max(D), arr.ind = TRUE) 
      last_point = three_points[!(three_points %in% rownames(farthest))] 

      ## If all the last three points have farthest distance 
      if (length(last_point) == 0) { 
        POINTS_to_use = rownames(Prev) 
        break 
      } 
      ## 
      cat("The RAD-Median is:", last_point, "\n")
      return(X[last_point,]) 
    } 
    #####

    farthest = which(D == max(D), arr.ind = TRUE) 
    D = D[!(rownames(D) %in% rownames(farthest)), !(colnames(D) %in% rownames(farthest))] 
    points = max(ncol(D), 0) 

    # Ensure that the number of points kept is the fewest possible while at least TRIM_Remain
    if (points == TRIM_Remain) { 
      POINTS_to_use = rownames(D) 
      break 
    } else if (points < TRIM_Remain) { 
      POINTS_to_use = rownames(Prev) 
      break 
    } 
  } 

  n_use = length(POINTS_to_use) 
  points_in_mean = rep(0, n_use) 
  for (a in 1:n_use) { 
    points_in_mean[a] = which(rownames(X) == POINTS_to_use[a]) 
  } 

  
  mean_of_points = colMeans(X[points_in_mean,]) 
  print("The points used are:") 
  print(X[points_in_mean,])
  
  if (RAD_Median == TRUE){
    print("The RAD-Median is:")
  } else{
    cat(paste("The Z", TRIM_percentage, "-Mean is:\n", sep = ""))
  }
  return(mean_of_points) 
}
```



# High-dimensional examples

Create an example dataset with 101 5-dimensional observations
```{r dataset_examples}
set.seed(1)
n = 101 # n is the number of observations
p = 5 # p is the number of coordinates of each observation
Sample_data = matrix(data = rnorm(n*p, mean = 0, sd = 10), nrow = n, ncol = p)
```

Compute the RAD-Median
```{r ex1}
TRIMMEDmean(Sample_data)
```

Compute the Z90-Mean
```{r ex2}
TRIMMEDmean(Sample_data, RAD_Median = FALSE, TRIM_percentage = 90)
```



# Example for 2-dimensional data, n odd

Compute the RAD-Median
```{r dataset_odd}
# Create the dataset
set.seed(12345) 
n = 35 # n is the number of observations
X <- matrix(data = runif(n = 2*n, min = 0, max = 10), nrow = n, ncol = 2, byrow = TRUE)

# Set row names as x1, x2, ...
rownames(X) <- paste0("x", 1:nrow(X))

# Set column names as c1, c2, ...
colnames(X) <- paste0("c", 1:ncol(X))

# Prepare data for plotting
plot_data <- as.data.frame(X)
plot_data$Point <- rownames(plot_data)  # Add point names for labels

TRIMMEDmean(X)
```

Plot the dataset and RAD-Median
```{r ex_odd}
# Compute the RAD-Median
trimmed_mean_result <- TRIMMEDmean(X)
print(trimmed_mean_result)

# Plot the points and RAD-Median
suppressWarnings({
  p <- ggplot(plot_data, aes(x = c1, y = c2)) +
    geom_point(color = "grey", size = 3) +  # Original points
    geom_point(aes(x = trimmed_mean_result[1], y = trimmed_mean_result[2]), 
               color = "red", shape = 5, size = 4, stroke = 1) +  
    xlab("Coordinate 1") +
    ylab("Coordinate 2") +
    theme_minimal(base_size = 15) +
    theme(
      axis.title.x = element_text(size = 14, face = "bold"),  # Larger x-axis title
      axis.title.y = element_text(size = 14, face = "bold"),  # Larger y-axis title
      axis.text.x = element_text(size = 12),  # Larger x-axis text
      axis.text.y = element_text(size = 12),  # Larger y-axis text
      axis.line = element_line(colour = "black"),  # Add axis lines
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      panel.border = element_blank()
    ) +
    #geom_text(aes(label = Point), hjust = 1.5, vjust = 0, size = 4) +
    coord_fixed(ratio = 1) +  # Fixed aspect ratio
    scale_x_continuous(expand = c(0, 0),
                       limits = c(0, 11)) +  # Set x-axis to start from 0
    scale_y_continuous(expand = c(0, 0),
                       limits = c(0, 11))   # Set y-axis to start from 0
    print(p)
})
```



# Example for 2-dimensional data, n even

Compute the RAD-Median
```{r dataset_even}
# Create the dataset
set.seed(2024) 
n = 40 # n is the number of observations
X <- matrix(data = runif(n = 2*n, min = 0, max = 10), nrow = n, ncol = 2, byrow = TRUE)

# Set row names as x1, x2, ...
rownames(X) <- paste0("x", 1:nrow(X))

# Set column names as c1, c2, ...
colnames(X) <- paste0("c", 1:ncol(X))

# Prepare data for plotting
plot_data <- as.data.frame(X)
plot_data$Point <- rownames(plot_data)  # Add point names for labels

TRIMMEDmean(X)
```
Plot the dataset and RAD-Median
```{r ex_even}
# Compute the RAD-Median
trimmed_mean_result <- TRIMMEDmean(X)
print(trimmed_mean_result)

# Plot the points and RAD-Median
suppressWarnings({
  p <- ggplot(plot_data, aes(x = c1, y = c2)) +
    geom_point(color = "grey", size = 3) +  # Original points
    geom_point(aes(x = trimmed_mean_result[1], y = trimmed_mean_result[2]), 
               color = "red", shape = 5, size = 4, stroke = 1) +  
    xlab("Coordinate 1") +
    ylab("Coordinate 2") +
    theme_minimal(base_size = 15) +
    theme(
      axis.title.x = element_text(size = 14, face = "bold"),  # Larger x-axis title
      axis.title.y = element_text(size = 14, face = "bold"),  # Larger y-axis title
      axis.text.x = element_text(size = 12),  # Larger x-axis text
      axis.text.y = element_text(size = 12),  # Larger y-axis text
      axis.line = element_line(colour = "black"),  # Add axis lines
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      panel.border = element_blank()
    ) +
    #geom_text(aes(label = Point), hjust = 1.5, vjust = 0, size = 4) +
    coord_fixed(ratio = 1) +  # Fixed aspect ratio
    scale_x_continuous(expand = c(0, 0),
                       limits = c(0, 11)) +  # Set x-axis to start from 0
    scale_y_continuous(expand = c(0, 0),
                       limits = c(0, 11))   # Set y-axis to start from 0
    print(p)
})
```



# Example for 2-dimensional data with outliers

Make a dataset with outliers
```{r dataset_outliers}
# Create the dataset
set.seed(2024) 
X <- matrix(data = c(rnorm(n = 30, mean = 25, sd = 5), runif(n = 160, min = 100, max = 150)), ncol = 2, byrow = TRUE)

# Set row names as x1, x2, ...
rownames(X) <- paste0("x", 1:nrow(X))

# Set column names as c1, c2, ...
colnames(X) <- paste0("c", 1:ncol(X))

# Prepare data for plotting
plot_data <- as.data.frame(X)
plot_data$Point <- rownames(plot_data)  # Add point names for labels
```

Compute the RAD-Median
```{r oulier_RAD}
TRIMMEDmean(X)
```

Plot the RAD_Median
```{r ex_outliers_RAD}
# Compute the RAD_Median
trimmed_mean_result <- TRIMMEDmean(X)
print(trimmed_mean_result)

# Plot the points and RAD_Median
suppressWarnings({
  p <- ggplot(plot_data, aes(x = c1, y = c2)) +
    geom_point(color = "grey", size = 3) +  # Original points
    geom_point(aes(x = trimmed_mean_result[1], y = trimmed_mean_result[2]), 
               color = "red", shape = 5, size = 4, stroke = 1) +  
    xlab("Coordinate 1") +
    ylab("Coordinate 2") +
    theme_minimal(base_size = 15) +
    theme(
      axis.title.x = element_text(size = 14, face = "bold"),  # Larger x-axis title
      axis.title.y = element_text(size = 14, face = "bold"),  # Larger y-axis title
      axis.text.x = element_text(size = 12),  # Larger x-axis text
      axis.text.y = element_text(size = 12),  # Larger y-axis text
      axis.line = element_line(colour = "black"),  # Add axis lines
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      panel.border = element_blank()
    ) +
    #geom_text(aes(label = Point), hjust = 1.5, vjust = 0, size = 4) +
    coord_fixed(ratio = 1) +  # Fixed aspect ratio
    scale_x_continuous(expand = c(0, 0),
                       limits = c(0, 160)) +  # Set x-axis to start from 0
    scale_y_continuous(expand = c(0, 0),
                       limits = c(0, 160))   # Set y-axis to start from 0
    print(p)
})
```

Compute the Z80-Mean
```{r oulier_Z80}
TRIMMEDmean(X, RAD_Median = FALSE, TRIM_percentage = 80)
```

Plot the Z80-Mean
```{r ex_outliers_Z80}
# Compute the Z80-Mean
trimmed_mean_result <- TRIMMEDmean(X, RAD_Median = FALSE, TRIM_percentage = 80)
print(trimmed_mean_result)

# Plot the points and Z80-Mean
suppressWarnings({
  p <- ggplot(plot_data, aes(x = c1, y = c2)) +
    geom_point(color = "grey", size = 3) +  # Original points
    geom_point(aes(x = trimmed_mean_result[1], y = trimmed_mean_result[2]), 
               color = "red", shape = 5, size = 4, stroke = 1) +  
    xlab("Coordinate 1") +
    ylab("Coordinate 2") +
    theme_minimal(base_size = 15) +
    theme(
      axis.title.x = element_text(size = 14, face = "bold"),  # Larger x-axis title
      axis.title.y = element_text(size = 14, face = "bold"),  # Larger y-axis title
      axis.text.x = element_text(size = 12),  # Larger x-axis text
      axis.text.y = element_text(size = 12),  # Larger y-axis text
      axis.line = element_line(colour = "black"),  # Add axis lines
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      panel.border = element_blank()
    ) +
    #geom_text(aes(label = Point), hjust = 1.5, vjust = 0, size = 4) +
    coord_fixed(ratio = 1) +  # Fixed aspect ratio
    scale_x_continuous(expand = c(0, 0),
                       limits = c(0, 160)) +  # Set x-axis to start from 0
    scale_y_continuous(expand = c(0, 0),
                       limits = c(0, 160))   # Set y-axis to start from 0
    print(p)
})
```